First, I would like to thank my advisor, Colin Dewey, for his kind and dedicated mentorship over the years and for teaching me how to conduct rigorous and insightful science.  I have immensely enjoyed these years learning from and working with him. His kindness and dedication has set for me a strong example of academic mentorship.  I would also like to thank my committee for their valuable guidance. I thank AnHai Doan for providing me direction in the early stages of the MetaSRA project in showing me how to iteratively develop the MetaSRA pipeline in a data-driven manner.  I wish to thank Mark Craven for not only serving on my committee, but also for teaching my first machine learning course during my first semester at UW-Madison, which helped to ignite my passion for this field. I thank Matt Ziegler for his help in building a fantastic interface for the MetaSRA website.  I am also indebted to the Computation and Informatics in Biology and Medicine training program for funding three years of my PhD and for providing me abundant opportunities to present my work. 

Earning a PhD is never an individual's accomplishment.  Rather, it requires a strong network of love and support, which I am extremely grateful to have had. First, I would like to thank my parents for their unconditional love and unwavering support throughout my entire life which made everything I have accomplished possible. To my brothers David and Max for their lifelong friendship, and those enjoyable late night conversations on topics both silly and serious. To Hannah for her love, for lifting me up during the hard times, and for pushing me to be the best person I can be.  To my Madison friends Shoban Chandrabose, Stephen Brown, Steve Lazzaro, Isaac Harrington, Marissa Gaskill, Shree Hardikar, Alisa Maas, Ross Kleiman, Alexi Brooks, and others who have helped me stay balanced - from board game nights to Comedy Club outings, nights on the Terrace to corn mazes. And lastly, a quick shout-out to the furry companionship of Duke, Korra, Taco, and Beeb.  Thank you all.